---
title: "Severe Weather Events Effects on US Population Health and Economy"
output: html_document
---

##Synopsis

In this report, we explore the U.S. National Oceanic and Atmospheric 
Administration's (NOAA) US storm database and aim to answer two questions: 
1) *Across the United States, which types of events (as indicated in the EVTYPE 
variable) are most harmful with respect to population health?* and 
2) *Across the United States, which types of events have the greatest economic 
consequences?* **NEED SUMMARY OF RESULTS HERE**

---

##Loading and Processing the Raw Data

The NOAA US storm database can be downloaded from the Coursera Data Science 
track site: 

- [https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) 

The data comes in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size.  
This database tracks characteristics of major storms and weather events in the 
United States, including when and where they occur, as well as estimates of any 
fatalities, injuries, and property damage.The following two supplementary 
resources provide information on how the data variables are constructed/defined:

- [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
- [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

###Download and read the data to a local file

The following code will first check the working directory to ensure the file has 
not already been downloaded. If the data file does not appear to exist in the 
working directly, the file will be downloaded (from the above location) and read
into the local cache.

```{r echo=TRUE, cache=TRUE}
fileName <- 'repdata-data-StormData.csv.bz2'
  fileExists <- fileName %in% dir(getwd())
  if (fileExists==FALSE){
    fileurl <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
    download.file(fileurl, destfile=fileName)
  }

theData <- read.csv(fileName)
```

---

##Results

Required R packages:
```{r echo=TRUE}
if (!require(dplyr)){ 
  install.packages('dplyr')
}
library(dplyr)
```

###*Question 1: Across the United States, which types of events are most harmful with respect to population health?*

To identify the most healthfully harmful events, we group the data by event type 
and summarize for fatalities and injuries.
Then the resulting dataframe is sorted from highest number of fatalities, and 
from highest number of injuries.
(Uses dplyr functions)

```{r echo=TRUE, cache=TRUE}
dat <- group_by(theData, EVTYPE)
eventHealth <- summarize(dat, totFatal=sum(FATALITIES), totInjured=sum(INJURIES))
eventHealth$EVTYPE <- as.character(eventHealth$EVTYPE)
eventFatalities <- arrange(eventHealth,desc(totFatal))
eventInjuries <- arrange(eventHealth,desc(totInjured))
```

Looking at the top 10 for the sorted list of totaled Fatalities and Injuries, we 
can see some matching event types.

```{r echo=TRUE}
##Sorted from highest number of fatalities
head(eventFatalities,10)
##Sorted from highest number of injuries
head(eventInjuries,10)
```

We can then compare the two sorted dataframes for the weather events that occur in both.


```{r echo=TRUE}
dat2 <- vector(mode="character", length=0)
for (i in 1:10){
  if (grepl(eventFatalities[i,1],eventInjuries[1:10,1])){
    dat2 <- c(dat2, as.character(eventFatalities[i, 1]))
  }
}
```

Based on the above logic, it appears that the weather events (as values within 
the variable EVTYPE) with the most harmful effects on US Population health are 
`r dat2`.

###*Question 2: Across the United States, which types of events have the greatest economic consequences?*

Economic consequences of a storm event will be measured by total property 
damage. Property damage is recorded in the dataset as a dollar amount within the 
PROPDMG variable and the magnitude of the dollar amount is recorded within the
PROPDMGEXP variable ('K' for thousands, 'M' for millions, and 'B' for billions 
of dollars in damage).

A new column is created in the dataset to provide a integer that correctly 
accounts for the magnitude of the dollar amount of damage.

```{r echo=TRUE}
datK <- filter(theData, PROPDMGEXP=='K')
datK <- mutate(datK, totDMG=PROPDMG*1000)
datM <- filter(theData, PROPDMGEXP=='M')
datM <- mutate(datM, totDMG=PROPDMG*1000000)
datB <- filter(theData, PROPDMGEXP=='B')
datB <- mutate(datB, totDMG=PROPDMG*1000000000)
dat0 <- filter(theData, PROPDMGEXP=='')
dat0 <- mutate(dat0, totDMG='0')
dataDMG <- rbind(datK, datM)
dataDMG <- rbind(dataDMG, datB)
dataDMG <- rbind(dataDMG, dat0)
```

Now summarize for the total damage for each event type.

```{r echo=TRUE}
datDMG <- group_by(dataDMG, EVTYPE)
datDMG$totDMG <- strtoi(datDMG$totDMG)
eventEconomies <- summarize(datDMG, totalDamage=sum(totDMG))

##Arrange the data to descend from most damange in dollars to least
eventEconomies <- arrange(eventEconomies, desc(totalDamage))
eventEconomies$EVTYPE <- as.character(eventEconomies$EVTYPE)
```

If we plot the top 10 event types by greatest number of damage in US dollars:

```{r echo=TRUE}
temp <- eventEconomies[1:10,]

```



